---
title: "Population projection with the popproj package"
author: Maria Osborne
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{popproj_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Summary

`popproj` is a package for conducting dynamic population projection.


## Set-up
```{r setup}
library(popproj)
library(data.table)
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
```



## Obtaining Vital Rates

As long as they are measured in one-year age intervals, vital rates may be sourced from anywhere or simulated. However, the package provides a shortcut for obtaining real rates from the Human Fertility Database and Human Mortality Database and formatting them for use in `project_leslie1()` or `project_leslie1_dynamic()`. Both `get_asfr_HFD()` and `get_pyl_HMD()` return a vector of length 111, corresponding to ages 0-110. One must be a registered user of the HFD and HMD with valid username and password to use these methods.

note: recent changes to the HMD and HFD websites have made these methods non-functional. to be fixed.
```{r}
# Fertility
#my_Fx <- get_asfr_HFD(country_code = 'NOR', year = 1980, hfd_username = "mariaosborne",
#             hfd_password = "")

# Mortality
#my_Lx <- get_pyl_HMD(country_code = 'NOR', year = 1980, hmd_username = "mariaosborne",
#             hmd_password = "") 
```


To demonstrate, we will use some pre-loaded data. The package contains age-specific fertility, mortality, and population data from Norway and Japan.


The `pad_fx_vector()` method will format the fertility data correctly, so that it has 111 age groups like the mortality data do.
```{r}
# ASFRs
my_Fx <- norway_asfr[Year == 1980]$ASFR
# pad the ASFR vector so that it will be the same length as the Lx vector
my_Fx <- pad_fx_vector(my_Fx)
my_Fx

# Lx values
my_Lx <- norway_female_lifetable[Year == 1980]$Lx
my_Lx <- (my_Lx)/100000
my_Lx
```




## Basic Projection

All population projection is based on leslie matrices with one-year age intervals. There are two methods 
for specifying a projection in this package. The first is `project_leslie1()`. This projection is based on a 
set of initial fertility and mortality rates, where life expectancy and NRR are allowed to vary over the
course of the projection.

- `nsteps`: the number of years to include in the projection

- `time`: can be used in place of nsteps. A vector of years to project for, e.g. 0:10 or -50:50, or 1960:2000.

- `nrr_vec`: vector of numbers specifying the net reproductive rate at each step in the 
projection. The length of `nrr_vec` must be `nsteps` or length(time)

- `e0_vec`: vector of numbers specifying life expectancy at each step in the projection. 
The length of `e0_vec` must be `nsteps` or length(time)

- `Lx_zero`: vector of initial L(x) values. One-year age groups

- `Fx_zero`: vector of initial F(x) values. Must be the same length at `Lx_zero` and have same age groups

- `starting_pop`: an optional vector of initial population by 1-year age groups. If left unspecified, projection will start from the stable age structure of the leslie matrix from `Lx_zero` and `Fx_zero`
normalized to a starting population of 1


As an example, we will perform a projection using data from Norway. In this scenario, we project for 120 years; NRR begins at 2 and falls to 1 at year 21 of the projection.
```{r}
# Example basic projection
# we can either specify the number of steps 
my_projection <- project_leslie1(nsteps=120,
                nrr_vec = c(rep(2,20), rep(1,100)),
                e0_vec = rep(80,120),
                Lx_zero = my_Lx,
                Fx_zero = my_Fx)
 
# ... or vector of years
# for example -10:10 or 1950:2000
my_projection_t <- project_leslie1(time = c(-20:99),
                nrr_vec = c(rep(2,20), rep(1,100)),
                e0_vec = rep(80,120),
                Lx_zero = my_Lx,
                Fx_zero = my_Fx)

```

The `project_leslie1` method returns a list, consisting of

-   `K_mat` : A matrix containing the result of the projection, where each row is an 
age group and each column is a step in the projection. The column sums are the population
at each year

-   `births_vec` : A vector of births over time

-   `nrr_vec`: A vector of NRR's over time

-   `e0_vec`: A vectors of e(0) values over time

-   `time_vec` A vector of years. If no time vector was provided, returns 0:(nsteps - 1)

```{r}
my_projection
```


Any returned elements of the list can be plotted with base R.
```{r}
# plot population size over time
plot(my_projection$time_vec, colSums(my_projection$K_mat), xlab = "year", ylab = "population")

# plot number of births
plot(my_projection$time_vec, my_projection$births_vec, xlab = "year", ylab = "births")

```

### prettier plotting
Built in plotting with the `plot_projection` method. Limits customization but easily enhances aesthetics.

Plots NRR, life expectancy, (female) population, and births over the course of the projection

```{r}
plot_projection(my_projection)
```




## Projections: Method 2

Alternatively, one can allow specify vital rates L(x) and F(x) at each step of the projection by using the 
`project_lesli1_dynamic()` method, which take the following arguments:

- `nsteps`: the number of steps in the projection

- `Lx_mat`: matrix of L(x) values for each step of the projection. Must have same number of columns as number of steps in the projection

- `Fx_mat`: matrix of F(x) values for each step of the projection. Must have same number of columns as number of steps in the projection and be the same dimensions as `Lx.mat`

- `starting_pop`: an optional vector of initial population by 1-year age groups. If left unspecified, projection will start from the stable age structure of the leslie matrix implied by initial L(x) and F(x) values


### Example with Norwegian data

As an example, we will project the Norwegian population 1970-2018 using real fertility/mortality rates & the actual population
of Norway in 1970. This should show us what would happened if the net migration rate in Norway dropped to zero after 1970.

(NOTE: not sure if things are starting at year 0 of year 1 here)

First, we will transform the data into a useful structure. Fx and Lx matrices can be built from
scratch, or can be gleaned from real HFD & HMD data using the `lifetable_to_matrix()` function.
```{r}
# Lx from lifetables
female_lifetable <- norway_female_lifetable

# ASFR data
fertility_rates <- norway_asfr

# convert to matrix
Lx_matrix <- lifetable_to_matrix(data = female_lifetable, years = 1970:2018, rate = 'Lx')
Fx_matrix <- lifetable_to_matrix(data = fertility_rates, years = 1970:2018, rate = 'ASFR')

Fx_matrix
(dim(Lx_matrix) == dim(Fx_matrix))
```


Now that the data are in matrices, let's get the starting population and project
```{r}
# Get starting population
population_by_year <- norway_population_size
starting_pop <- population_by_year[Year == 1970, Female]

# dynamic projection
norway_dynamic <- project_leslie1_dynamic(time = 1970:2018,
                 Lx_mat = Lx_matrix,
                 Fx_mat = Fx_matrix,
                 starting_pop = starting_pop)

# look at population over time
plot(colSums(norway_dynamic$K_mat), xlab = "year", ylab = "population", main = "Projected")

# investigate other elements
#norway_dynamic$nrr_vec
#norway_dynamic$births_vec
#norway_dynamic$e0_vec

# compare the projected population to real population growth
real_population <- population_by_year[Year %in% 1970:2018, .(fpop = sum(Female)), by = Year]
million <- 1000000
plot(norway_dynamic$time_vec, colSums(norway_dynamic$K_mat), ylim = c(1.9*million, 2.65*million),
     ylab = "female population", xlab = "year", main = "Projection vs. Actual")
lines(norway_dynamic$time_vec, real_population$fpop)


# notes to self
# nrr look reasonable?
# no peaks
```

The actual female population of Norway in 2018 is 2.63 million. If Norway had become a closed population in 
1970, the projection estimates a female population of about 2.20 million.


### Example with Japanese data
Net migration in Japan is very low, so using the dynamic projection should predict very similar numbers as the real population. Let's make sure things are working ok by checking that...
```{r}
# Lx 
japan_female_lifetable <- japan_female_lifetable 
# Fx
japan_Fx <- japan_asfr 

# convert to matrices
japan_Lx_mat <- lifetable_to_matrix(data = japan_female_lifetable, years = 1960:2010, rate = 'Lx')
japan_Fx_mat <- lifetable_to_matrix(data = japan_Fx, years = 1960:2010, rate = 'ASFR')

# population
japan_Nx <- japan_population_size
japan_initial_pop <- japan_Nx[Year == 1960,]$Female
japan_real_population <- japan_Nx[Year %in% 1960:2010, .(fpop = sum(Female)), by = Year]


# project
japan_dynamic <- project_leslie1_dynamic(time = 1960:2010,
                 Lx_mat = japan_Lx_mat,
                 Fx_mat = japan_Fx_mat,
                 starting_pop = japan_initial_pop)
```


Let see how the projection (points) stacks up against the actual female population (lines)
```{r}
plot(japan_dynamic$time_vec, colSums(japan_dynamic$K_mat), pch = 20,
     xlab = "year", ylab = "population")
lines(japan_real_population)
```
The projection tracks very closely.
projected 2010 female pop: 64,524,828
real 2010 female pop: 64,748,265

lastly, let's look at all plots for the Japan projection:
```{r}
plot_projection(japan_dynamic)
```




## Coale scenario simulation
Projecting exponentially declining fertility with constant mortality

`coale_projection()` method is meant to replicate the peak projection shiny app.
So every projection is done for times -50:50, where NRR hits 1 at year 0
```{r}
# default of NRR declining by 1.5% per year, e0 of 80
# we will use initial Lx and Fx values from the Norwegian data
coale <- coale_projection(Lx_zero = Lx_matrix[,1],
                          Fx_zero = Fx_matrix[,1],
                          k1 = -0.015,   # rate of NRR decline
                          k2 =  0.0001, # extra curvature (optional)
                          e0_start = 80) # life expectancy at time 0

plot_projection(coale)

# rho parameter problematic

#plot(colSums(coale$K_mat))
#plot(coale$births_vec)
#plot(coale$e0_vec)
#plot(coale$time_vec, coale$nrr_vec)

# notes
# scaling arg?
```



## Predicted Peaks

We can use the `plot_projection()` method again with additional options: 

1. Predicted Peak Values: Plot vertical lines where births and population are predicted to reach
their maximum

2. Generational Slopes: Draw a chord between two generations on the birth and population plots.
Default generation length is 30 years. 


```{r}
plot_projection(coale, plot_predicted_peaks = T, 
                plot_generational_slopes = T, 
                generation_t = 0)

# predicted peaks can be inaccurate if there are multiple local maxima
```







