---
title: "Population projection with the popproj package"
author: Maria Osborne
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{popproj_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Summary

`popproj` is a package for conducting population projection.


## Set-up
```{r setup}
library(popproj)
library(data.table)
library(dplyr)
library(tidyr)
```



## Obtaining Vital Rates

As long as they are measured in one-year age intervals, vital rates may be sourced from anywhere or simulated. However, the package provides a shortcut for obtaining real rates from the Human Fertility Database and Human Mortality Database and formatting them for use in `project_leslie1()` or
`project_leslie1_dynamic()`. Both `get_asfr_HFD()` and `get_pyl_HMD()` return a vector of length 111, corresponding to ages 0-110. One must be a registered user of the HFD and HMD with valid username and password to use these methods.

note: get_asfr_HFD is currently not working due to updates in the HFD website
```{r}
# Fertility
#my_Fx <- get_asfr_HFD(country_code = 'NOR', year = 1980, hfd_username = "--",
#             hfd_password = ""--")

# Mortality
#my_Lx <- get_pyl_HMD(country_code = 'NOR', year = 1980, hmd_username = "--",
#             hmd_password = "--") 
```


in order to not put my HMD/HFD passwords online, I will read in some pre-downloaded data
```{r}
# ASFRs
my_Fx <- fread("/hdir/0/mariaosborne/censoc/popproj/NORasfrRR.txt")[Year == 1980]$ASFR
# pad the ASFR vector so that it will be the same length as the Lx vector
my_Fx <- pad_fx_vector(my_Fx)
my_Fx

# Lx values
my_Lx <- fread("/hdir/0/mariaosborne/censoc/popproj/norway_flt.txt")$Lx
# normalize to radix of 1
my_Lx <- (my_Lx)/100000
my_Lx
```




## Basic Projection

All population projection is based on leslie matrices with one-year age intervals. There are two basic ways 
to set up a projection in this package. The first is `project_leslie1()`. This projection is based on a 
set of initial fertility and mortality rates, where life expectancy and NRR are allowed to vary over the
course of the projection.

- `nsteps`: the number of steps to project for

- `nrr.vec`: vector of numbers specifying the net reproductive rate at each step in the 
projection. The length of `nrr.vec` must be `nsteps`.

- `e0.vec`: vector of numbers specifying life expectancy at each step in the projection. 
The length of `e0.vec` must be `nsteps`.

- `Lx.zero`: vector of initial L(x) values. 

- `Fx.zero`: vector of initial F(x) values. Must be the same length at `Lx.zero`

- `starting.pop`: an optional vector of initial population by 1-year age groups. If left unspecified, projection will start from the stable age structure of the leslie matrix from `Lx.zero` and `Fx.zero`
normalized to a radix of 1


As an example, we will perform a projection using data from Norway. In this scenario, we project for 100 years; NRR begins at 2 and falls to 1 at year 21 of the projection.
```{r}
# Example basic projection
# we can either specify the number of steps 
my_projection <- project_leslie1(nsteps=120,
                nrr.vec = c(rep(2,20), rep(0.9,100)),
                e0.vec = rep(80,120),
                Lx.zero = my_Lx,
                Fx.zero = my_Fx)
 
# ... or a time vector. must be in one year increments
my_projection_t <- project_leslie1(time = c(-20:99),
                nrr.vec = c(rep(2,20), rep(0.9,100)),
                e0.vec = rep(80,120),
                Lx.zero = my_Lx,
                Fx.zero = my_Fx)
my_projection
```

The `project_leslie1` method returns a list, consisting of

-   `K_mat` : A matrix containing the result of the projection, where each row is an 
age group and each column is a step in the projection.

-   `births_vec` : A vector of births over time

-   `nrr_vec`: A vector of NRR's over time

-   `e0_vec`: A vectors of e(0) values over time


Any returned elements can be plotted with base R. We will do more advanced/pretty plotting later
```{r}
# plot population size over time
plot(colSums(my_projection$K_mat))  

# plot number of births
plot(my_projection$births_vec)

#plot_projection(my_projection_t, plot_predicted_peaks = T)
# non differentiable points?
#get_approx_peak(t=my_projection$time_vec, ft=my_projection$births_vec)
  #dft <- center_diff(ft)
  #peak <- approx(x = dft, y = t, xout = 0)$y
#center_diff(my_projection_t$births_vec) 
```


Alternatively, one can allow specify vital rates L(x) and F(x) at each step of the projection by using the 
`project_lesli1_dynamic()` method, which take the following arguments:

- `nsteps`: the number of steps in the projection

- `Lx.mat`: matrix of L(x) values for each step of the projection. Must have same number of columns as number of steps in the projection

- `Fx.mat`: matrix of F(x) values for each step of the projection. Must have same number of columns as number of steps in the projection and be the same dimensions as `Lx.mat`

- `starting.pop`: an optional vector of initial population by 1-year age groups. If left unspecified, projection will start from the stable age structure of the leslie matrix from `Lx.zero` and `Fx.zero`


As an example, we will project using real fertility and mortality rates from Norway,
And starting population is Norway in 1970. This will be similar to how the population
of Norway would have changed if net migration had dropped to zero after 1970.

(NOTE: need to check the number of steps/dimensions at some points -- 
not sure if starting at 0 or 1)

First, we will transform into a useful structure. Fx and Lx matrices can be built from
scratch, or can be gleaned from real HFD & HMD data using the `lifetable_to_matrix()` function.
```{r}
# Matrix of Lx values
female_lifetable <- fread("/hdir/0/mariaosborne/censoc/popproj/norway_flt_allyears.txt")
fertility_rates <- fread("/hdir/0/mariaosborne/censoc/popproj/NORasfrRR.txt")

# convert to matrix
Lx_matrix <- lifetable_to_matrix(data = female_lifetable, years = 1970:2018, rate = 'Lx')
Fx_matrix <- lifetable_to_matrix(data = fertility_rates, years = 1970:2018, rate = 'ASFR')

(dim(Lx_matrix) == dim(Fx_matrix))
```


And perform the projection
```{r}
# Get starting population
population_by_year <- fread("/hdir/0/mariaosborne/censoc/popproj/norway_popsize.txt")
starting_pop <- population_by_year[Year == 1970, Female]

# an example here
my_dynamic_projection <- project_leslie1_dynamic(nsteps=49,
                 Lx_mat = Lx_matrix,
                 Fx_mat = Fx_matrix,
                 starting.pop = starting_pop)

# look at population over time
plot(colSums(my_dynamic_projection$K.mat))

# investigate other elements
my_dynamic_projection$nrr.vec
my_dynamic_projection$births.vec
my_dynamic_projection$e0.vec

# notes to self
# how does this nrr stack up against real data?
# it looks pretty good (a little less than 1/2 of the TFR), but need to check
# more thoroughly
# ^ dunno what this comment means anymore?
# I should try the plot_projection method on this one too, see if any issues
# no peaks
```


## Coale(?) simulation
Specific scenario of exponentially declining fertility with constant mortality
meant to replicate the shiny app
```{r}
# default of NRR declining by 1.5% per year, e0 of 80
coale <- coale_projection(Lx_zero = Lx_matrix[,1], Fx_zero = Fx_matrix[,1], k2 = 0.00015)

plot(colSums(coale$K_mat))
plot(coale$births_vec)
plot(coale$e0_vec)
plot(coale$nrr_vec)

#coale$nrr_vec
#coale$births_vec
#coale$e0_vec

# notes
# need to fix variable names!! e0_vec vs. e0.vec
#time goes from negative to positive
#(plot year zero)
#we should all a boolean argument for scaling into this method.
```


## Plotting Results
Each project method returns a list, the plot_projection() method plots all four
elements in a visually appealing way with additional options


### Basic Plots

By default, this method plots population size, births, NRR, and e0.

```{r}
plot_projection(coale)
```


### Additional Options
`plot_projection()` has additional options for:

1. NRR replacement: If there is a point where NRR reaches 1, mark it on the NRR graph with a vertical
line (not implemented yet; this is always plotted)

2. Generational Slopes: I think manually set either a starting or ending point
have the defauly generational length be 30, but alterable
throw and exception 

3. Predicted Peak Values: Plot vertical lines where births and population are predicted to reach
their maximum

```{r}
plot_projection(coale, plot_predicted_peaks = T, 
                plot_generational_slopes = T, generation_t = 100)
```













